High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2020-11-10 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the {\tt uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$3 times finer than natural weights. The model image is obtained by fitting the data using Chi-squared minimisation, with a measure of regularization when required. Here we adopted a $\chi^2$ model image with $\lambda$ = 0.01 for regularization (***mem\_lS0.01\_lL0.0\_nogrid/mod\_out.fits). The resulting {\tt uvmem} image shown in Fig.~\ref{fig:images_vs_simulated} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, beyond which it drops steeply before breaking at $\sim$35\,au into a shallow tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} did anticipate its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au.

@@

Repeating the optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 both share the same orientation and centre, given the errors, and both are offset from the star. Even though all orientation parameters for Ring13 and Ring24 are consistent within the errors, there are however some hints for a different orientation, as summarised in Fig.~\ref{fig:polarring}, perhaps due to the joint effect of all these small differences.

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{o}(r) \, [r/r_{o}(r)]^{\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at r = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. For the gas and the small dust grains $\chi$ = 1 and for the millimetre-size dust population we assign a fixed $\chi$ = 1/2 for simplicity, as in \citet{Rosenfeld_2013}.

For the vertical structure, \citet{dOrazi} found flaring angles of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model uses the same values, with two different flaring indexes $\psi$. The separation between the two values were set at $r$ = 19\,au with $\psi=0.2$ for Ring6 and Ring13, and $\psi=0.5$ for Ring24. The scale height is $H_\circ = 0.045$\,au at $r_\circ = 19.5$\,au.

@@

The inner radius of the model grid was set to 0.1\,au and an outer radius of 115\,au, large enough for the dust disc to become undetectable. We set the values of the inclination an the position angle as the same as those obtained from the ALMA observation in Section~\ref{sec:Observations}, so the model has an inclination of i = 32.96\,$\degr$ and a P.A. = 77.31\,$\degr$. Finally, the distance is set at d = 72.4\,pc \citep{Gaia}.

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a good approximation of the disc structure, including the dimensions of Ring13. Accordingly, taking the parametric model values for Ring13 we have that it has a radius of 13.35\,au, a width of 2.1\,au, and a scale height FWHM of the millimetre-size dust ranging from 0.30 to 0.36\,au. The total dust mass of Ring13 is about 3\,M$_{\earth}$ in the model***. In turn, in our model Ring24 reaches peak intensity at $\sim$30\,au, then breaks at $\sim$35\,au, as in the observations, and has a total dust mass of $\sim$43\,M$_{\earth}$. The model predictions for the millimetre-size dust population in Ring13 are close to the measurements, with only a 0.03\,au difference in the centroid location of the ring, and a 0.6\,au difference in the width estimation.
%As \citet{2018ApJ...869L..46D} explain the stability of a narrow dust ring requires that its height is smaller than its radial extent. 
Using the parametric model scale height of 0.33\,au at 13.32\,au and the measured width of Ring13 (1.50\,au FWHM, see Section~\ref{sec:Observations}), we have that its width is $\sim$4.6 times its scale height.

The observed structure may point to the existence of planet-disk interaction within this system, where a giant planet depletes its orbit of gas and dust material. A possible constellation in this scenario is therefore the presence of two giant planets in the disk, one planet between the star and Ring13, and one planet between Ring13 and Ring24. As \citet{Ru_z_Rodr_guez_2019} suggests, the planet between Ring13 and Ring24 may be a Jovian-mass planet with a mass within the range of 0.3$-$1.5\,$\mathrm{M}_{\mathrm{Jup}}$. This idea is supported by the similarity to the structure around HD 169142 which \citet{bertrang_avenhaus_2018} reproduced by including several giant planets. We point out that also the disk dispersal by photoevaporative winds \citep{1994ApJ...428..654H} as causation of the inner cavity cannot be ruled out at his stage.

@@

Similarly to \citet{2018ApJ...869L..46D}, the ratio between dust and gas ring width can be linked to important properties:
\begin{equation}\label{eq:ringwidth}
        \frac{w_\mathrm{d}}{w_\mathrm{g}} = \sqrt{\frac{\alpha}{\alpha+\mathrm{St}}}\,,
\end{equation}
where we implicitly assumed the turbulent diffusivity to be equal to the turbulent viscosity. Here, $\alpha$ represents the typical parameter for turbulent viscosity, and $\mathrm{St}$ is the dimensionless Stokes number. Assuming that the most relevant grain size for observed fluxes corresponds to $a = \lambda_{\mathrm{obs}}/(2 \pi)$, and estimating the gas surface density within Ring13 from the RT model to be roughly $20\,\mathrm{g}/\mathrm{cm}^2$, yields $\mathrm{St}\approx 0.03$. With these considerations, equation~\ref{eq:ringwidth} provides a value of $\alpha \approx 0.001$ for the level of turbulent viscosity. We emphasise that this relies on rough estimates and is additionally restricted to the area within Ring13.

>>>>

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the {\tt uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$3 times finer than natural weights. The model image is obtained by fitting the data using Chi-squared minimisation, with a measure of regularization when required. Here we adopted a $\chi^2$ model image with $\lambda$ = 0.01 for regularization (***mem\_lS0.01\_lL0.0\_nogrid/mod\_out.fits**). The resulting {\tt uvmem} image shown in Fig.~\ref{fig:images_vs_simulated} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, beyond which it drops steeply before breaking at $\sim$35\,au into a shallow tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} did anticipate its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au.

@@

Repeating the optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 both share a very similar orientation and centre, given the errors, and both are offset from the star. %Even though all orientation parameters for Ring13 and Ring24 are consistent within the errors, 
Perhaps due to the joint effect of all these small differences, we can see that there are some hints for a different orientation, as summarised in Fig.~\ref{fig:polarring}, .

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{o} \, [r/r_{o}]^{\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at r = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. For the gas and the small dust grains $\chi$ = 1 and for the millimetre-size dust population we assign a fixed $\chi$ = 1/2 for simplicity, as in \citet{Rosenfeld_2013}.

For the vertical structure, \citet{dOrazi} found flaring angles of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model uses the same values, with two different flaring indexes $\psi$. The separation between the two values were set at $r$ = 19\,au with $\psi=0.2$ for Ring6 and Ring13, and $\psi=0.5$ for Ring24. The scale height is $H_\circ = 0.8775$\,au at $r_\circ = 19.5$\,au.

@@

The inner radius of the model grid was set to 0.1\,au and an outer radius of 100\,au, large enough for the dust disc to become undetectable. We set the values of the inclination an the position angle as the same as those obtained from the ALMA observation in Section~\ref{sec:Observations}, so the model has an inclination of i = 32.96\,$\degr$ and a P.A. = 77.31\,$\degr$. Finally, the distance is set at d = 72.4\,pc \citep{Gaia}.

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a good approximation of the disc structure, including the dimensions of Ring13. Accordingly, taking the parametric model values for Ring13 we have that it has a radius of 13.35\,au, a width of 2.1\,au, and a scale height FWHM of the millimetre-size dust ranging from 0.30 to 0.36\,au. The total dust mass of Ring13 is about 1\,M$_{\earth}$ in the model. In turn, in our model Ring24 reaches peak intensity at $\sim$30\,au, then breaks at $\sim$35\,au, as in the observations, and has a total dust mass of $\sim$32\,M$_{\earth}$. The model predictions for the millimetre-size dust population in Ring13 are close to the measurements, with only a 0.03\,au difference in the centroid location of the ring, and a 0.6\,au difference in the width estimation.
%As \citet{2018ApJ...869L..46D} explain the stability of a narrow dust ring requires that its height is smaller than its radial extent. 
Using the large grains scale height of 0.65\,au at Ring13, and a measured width of 1.50\,au, the ring is $\sim$2.3 times more extended in radial than in vertical direction.

The observed structure may point to the existence of planet-disk interaction within this system, where a giant planet depletes its orbit of gas and dust material. A possible constellation in this scenario is therefore the presence of two giant planets in the disk, one planet between the star and Ring13, and one planet between Ring13 and Ring24. As \citet{Ru_z_Rodr_guez_2019} suggest, the putative planet between Ring13 and Ring24 may be a giant planet with a mass within the range of 0.3$-$1.5\,$\mathrm{M}_{\mathrm{Jup}}$. This idea is supported by the similarity to the structure around HD 169142 which \citet{bertrang_avenhaus_2018} reproduced by including several giant planets. %We point out that also the disk dispersal by photoevaporative winds \citep{1994ApJ...428..654H} as causation of the inner cavity cannot be ruled out at his stage.

@@

Similarly to \citet{2018ApJ...869L..46D}, the ratio between dust and gas ring width can be linked to important properties:
\begin{equation}\label{eq:ringwidth}
        \frac{w_\mathrm{d}}{w_\mathrm{g}} = \sqrt{\frac{\alpha}{\alpha+\mathrm{St}}}\,,
\end{equation}
where we implicitly assumed the turbulent diffusivity to be equal to the turbulent viscosity. Here, $\alpha$ represents the typical parameter for turbulent viscosity, and $\mathrm{St}$ is the dimensionless Stokes number. Assuming that the most relevant grain size for observed fluxes corresponds to $a = \lambda_{\mathrm{obs}}/(2 \pi)$, and estimating the gas surface density within Ring13 from the RT model to be roughly $20\,\mathrm{g}/\mathrm{cm}^2$, yields $\mathrm{St}\approx 0.03$. With these considerations, equation~\ref{eq:ringwidth} provides a value of $\alpha \approx 0.001$ for the level of turbulent viscosity. We emphasise that this relies on rough estimates and is additionally restricted to the area within Ring13. Yet, its moderate value agrees with the limit \citet{2018ApJ...869L..46D} inferred from multiple dust rings at larger orbital distances ($\alpha \gtrsim 5\times10^{-4}$). Further, it is also consistent with the non-detection of non-thermal broadening of CO lines reported by \citet{Flaherty_2020}, from which the authors deduced an upper limit of $\alpha \lesssim 10^{-3}$ at the midplane of V4046\,Sgr.
